---
title: "0. Data"
author: "Matteo Vecchi"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output: 
  workflowr::wflow_html:
    code_folding: hide 
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#
```{r packages, message = FALSE}
library(tidyverse)
library(dplyr)
library(readxl)  # To read from excel files
library(xfun)    # Download file from html report
```

This document describes the data analysis and plotting 

## Freezing experimental data

The original data file is `data_freezing.xlsx` and can be downloaded from here:
```{r echo=FALSE}
embed_file('./data/data_freezing.xlsx')
```
.  
  
The excel spreadsheet contains 4 sheets:

* They are:
  + **data_cycles**:  Experimental data on repeated daily freezing cycles survival. The columns are:
    + *ID*: the individual tardigrades ID.
    + *0* to *7*: each freezing-thaw cycle (1 = alive / 0 = dead).
    + *species_code*: species code.
    + *unfrozen*: if in any of the freezing cycles the water in the well didn´t froze when placed at -7C.
    + *when*: for unfrozen = 1, at which cycle the water didn´t freeze.
  + **data_longterm**:  Experimental data on one long term (one month) freezing survival.The columns are:
    + *species_code*: species code.
    + *survival*: number of animals alive after 1 month of freezing.
    + *total*: total number of animals used in the experiment.
  + **species_data**:  Data on the species/populations used in the experiments, including environmental data from their sampling sites. The columns are:
    + *species_code*: species code.
    + *Lat* and *Lon*: origin coordinates of the used populations.
    + *Origin*: origin of the used populations.
    + *others*: **OTHERS TO UDATE!!!!**
  + **species_abbrev**:  Species abbreviations used in the other sheets. The columns are:
    + *species_code*: species code.
    + *species*: species name.
    
## Phylogenetic tree
A phylogenetic tree of the used species has been obtained with MrBayes with the concatenated alignment of the markers 18S, 28S and *COI*.  
The tree can be downloaded here: 
```{r echo=FALSE}
embed_file('./data/tree.nex')
```
