---
title: "2. Long term freezing vs. freezing cycles"
author: "Matteo Vecchi"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output: 
  workflowr::wflow_html:
    code_folding: hide 
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message = FALSE}
library(tidyverse)
library(dplyr)
library(readxl)  # To read from excel files
library(R2jags)
library(ggplot2)
library(patchwork)
library(DT)
library(xfun)          # Download file from html report
library(ape)
library(phytools)
library(svglite)
```

We will test if there is a relationship between the resistence to freeze-thaw cycles and the resistence to long term freezing. Due to the low number of species tested, and the presence of a phylogenetic structure in both variables, we will use the simplest possible way that is testing the relationship between Phylogenetic Indipendent Contrasts (PIC).  

## Data preparation
```{r echo=T, results='hide', warning=FALSE,message=FALSE,error=FALSE}
# Read the data on long-term freezing sirvival and merge it in a single table with the estimated number of cycles to death of 50% of individuals for each tested species.
data_comparison_freezing = read_xlsx("./data/data_freezing.xlsx", sheet = "data_longterm") %>% mutate(p_long = as.numeric(survival)/as.numeric(total)) %>% merge(read.table("./output/means_cycles.txt", header=T)) %>% column_to_rownames(var = "species_code") %>% filter(complete.cases(.))
rownames(data_comparison_freezing) = data_comparison_freezing$species
```

## Phylogenetic tree loading and ultrametrization
```{r echo=T, message = FALSE, warning = FALSE, results='hide'}
# Load the species phylogenetic tree and ultrametricize it. As we are using the tree only to control for phylogeny, we don't need to get very fancy with ultrametrization and we can simply ise the "chronos" function.
tree = chronos(read.nexus("./data/tree.nex"))

# To avoid issues downstream, we crosscheck that all the tree tips are represented in the data table and viceversa
tree = drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% rownames(data_comparison_freezing))])

data_comparison_freezing = data_comparison_freezing[rownames(data_comparison_freezing) %in% tree$tip.label,]

```

## Ancestral state reconstruction of freezing tolerance
```{r echo=T, results='hide'}
# We use the "contMap" function from "phytools" to reconstruct the ancestral state.
vect.M = data_comparison_freezing$M
names(vect.M) = rownames(data_comparison_freezing)
objM = contMap(tree,vect.M,plot=FALSE, method="anc.ML", lims = c(0,30))

vect.p = data_comparison_freezing$p_long
names(vect.p) = rownames(data_comparison_freezing)
objp = contMap(tree,vect.p,plot=FALSE, method="anc.ML", lims = c(0,1))

```

## Ancestral state plotting
```{r,results='hide',fig.keep='all'}
par(mfrow=c(1,2))
plot(objM,lwd=7,ftype="reg",xlim=c(-0.2,1.2),legend=0.5, fsize=0.75)
plot(objp,lwd=7,direction="leftwards",ftype="off",xlim=c(-0.2,1.2),
    legend=0.5)
    
svglite("./output/M_vs_plong_plot.svg", width = 6, height = 6) # save the plot
par(mfrow=c(1,2))
plot(objM,lwd=7,ftype="reg",xlim=c(-0.2,1.2),legend=0.5)
plot(objp,lwd=7,direction="leftwards",ftype="off",xlim=c(-0.2,1.2),
    legend=0.5)
dev.off()
```

Download the plot in svg format
```{r echo=FALSE}
embed_file('./output/M_vs_plong_plot.svg')
```

## Correlation between long-term and repeated freezing
Test the correlation (corrected by phylogeny) between freeze-thaw cycles tolerance and long term freezing tolerance
```{r echo=T}
mat.chars = data_comparison_freezing[,c(3,5)] %>% drop_na() %>% mutate(p_long = scale(p_long), M = scale(M)) %>% as.matrix()

## calculate phylogenetic indipendent contrasts
M_pic = pic(mat.chars[,1],tree)
plong_pic = pic(mat.chars[,2],tree)

lm_pic = lm(M_pic ~ plong_pic -1) # Regression by the origin
plot(M_pic ~ plong_pic, pch=19,
     xlab="Long term freezing resistance PIC",
            ylab="Freeze-thaw resistance PIC")

a_pic = anova(lm_pic)
a_pic
``` 

The p-value of the correlation between tolerance to freeze-thaw cycles and long term survival (accounting for shared evolutionary history) is `r round(a_pic[[5]][1],3)`. This would suggest that there is no correlation between the two types of freezing resistances, however the sample size is very low (9 species,8 PIC) and by checking the scatterplot above it seems that only one point is responsible for this lack of correlation, so this result should be taken with extreme cation.
