---
title: "3. Freezing cycles tolerance vs. environmental variables"
author: "Matteo Vecchi"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output: 
  workflowr::wflow_html:
    code_folding: hide 
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(rgdal)
library(raster)
library(ape)
library(bayestestR)
library(R2jags)
library(patchwork)
library(DT)
library(ggmcmc)
library(xfun)
```

## Load sampling coordinates of used species
```{r load points coordinates}
points = data.frame(read_xlsx("./data/data_freezing.xlsx", sheet = "species_data"))[,1:3]
```



## Get climatic data for species sampling locations

```{r, eval = FALSE}
# This script extracts how many months have the minimum temperature below zero and the maximum temperature above zero (from Bioclim rasters) and save it in a table. 
# As the raw climatic rasters are very heavy (~ 10Gb), this code is not run, but only shown here.
# The tmin and tmax data rasters can be downloaded at: https://www.worldclim.org/data/monthlywth.html


tmin.list = list.files(path="./data/tmin", 
                           pattern =".tif", full.names=TRUE)
tmin.stack = raster::stack(x=tmin.list)
tmins = data.frame(raster::extract(tmin.stack, points[,3:2]))
tmins = cbind(points$species_code, tmins)
tmins = gather(tmins, "month", "temp_min", 2:ncol(tmins))
colnames(tmins)[1] ="species"
tmins$month = substr(tmins$month,start=17, stop = 23)


tmax.list = list.files(path="./data/tmax", 
                       pattern =".tif", full.names=TRUE)
tmax.stack = raster::stack(tmax.list)
tmax = data.frame(raster::extract(tmax.stack, points[,3:2]))
tmax = cbind(points$species_code, tmax)
tmax = gather(tmax, "month", "temp_max", 2:ncol(tmax))
colnames(tmax)[1] ="species"
tmax$month = substr(tmax$month,start=17, stop = 23)


temp_all = distinct(merge(tmins, tmax, by=c("species","month"))) 


  low = (temp_all$temp_min<0) * (temp_all$temp_max<0) # not really necessary
  mixed = (temp_all$temp_min<0) * (temp_all$temp_max>0)
  high = (temp_all$temp_min>0) * (temp_all$temp_max>0)

month_class = rep("low", nrow(temp_all))
month_class[as.logical(mixed)] = "mixed"
month_class[as.logical(high)] = "high"
  
temp_all$month_class = month_class

sum_table = data.frame(table(temp_all$species, temp_all$month_class))
colnames(sum_table) = c("species_code", "type", "count")
sum_table = sum_table %>% filter(type == "mixed") %>% mutate(prop_mixed = count/108)

write.table(sum_table, "./data/mixedmonths_data.txt")

```


```{r load proportion of mixed months}
bioclims_extracted = read.table("./data/mixedmonths_data.txt", header=T)

```

## Merge climatic data with freezing cycles tolerance data
```{r merge environmental data with freezing tolerance data}
data_merged = merge(read.table("./output/means_cycles.txt", header=T),bioclims_extracted)
data_merged$scaled_M = as.numeric(scale(data_merged$M))
```

## Load phylogenetic tree
```{r load phylogenetic tree, , results='hide'}
tree = chronos(read.nexus("./data/tree.nex"))
tree = drop.tip(tree, tip = tree$tip.label[!(tree$tip.label %in% data_merged$species)])
class(tree) = "phylo"

data_merged = data_merged[data_merged$species %in% tree$tip.label,]
```

## Run JAGS model
```{r jags model, results='hide'}

phylo.matrix = vcv.phylo(tree)
inv.phylo.matrix = solve(phylo.matrix)

data_merged = data_merged[match(rownames(inv.phylo.matrix),data_merged$species),]

data.jags = list(M = data_merged$scaled_M,
                 mixprop = as.numeric(scale(data_merged$prop_mixed)),
                 inv.phylo.matix = inv.phylo.matrix,
                 nsp = nrow(inv.phylo.matrix),
                 zeros = rep(0, nrow(inv.phylo.matrix)))

# The model includes some truncated priors. As it is not possible to do it by specifying the modell as function in R, the moded is loaded as txt file.

parameters.jags = c("alpha","beta","sigma.phylo","sigma.res")


ifelse("mod.env.rds" %in% list.files("./output"),{fit.env = readRDS("./output/mod.env.rds")},{

  fit.env = jags(data = data.jags,
               parameters.to.save = parameters.jags,
               model.file = "./code/model.txt",
               n.chains = 3, n.iter = 1000000)
write_rds(fit.env, file="./output/mod.env.rds")
  
})

# Make diagnostic plots (not showed but saved as pdf)
ggmcmc(ggs(as.mcmc(fit.env)), file="./output/model_fit.env.pdf", param_page=2)
```

Download model file here:
```{r echo=FALSE}
embed_file('./output/mod.env.rds')
```
  
Download diagnostic plots here:
```{r echo=FALSE}
embed_file('./output/model_fit.env.pdf')
```


## Calculate p.values and show model estimates
```{r get model bayesian pvalues}
chains = data.frame(do.call(rbind,as.mcmc(fit.env)))

p_dir = p_direction(chains)
p_vals = pd_to_p(p_dir$pd)
p_vals[3:5] = rep(NA,3)
names(p_vals) = p_dir$Parameter

fit.env$BUGSoutput$summary %>% cbind(p_vals) %>% round(4) %>%datatable(class = 'cell-border stripe')

```

The beta paramenter represent the effect of the proportion on "mixed" months on the resistance to freeze-thaw cycles. It is positive and significant, so the more the mixed months there are, the more the tardigrades are resistent.  

